Positivity-Preserving Adaptive Runge-Kutta Methods

### 2020-03-09 latex/overview.tex
Multiple small corrections
overview.tex:
spelling mistakes and minor notation changes
Changed the second example to make it more reasonable
TODOs: Abstract, Introduction, Formulation LPs, Numerical results, Conclussion

Bib_RK_optim.bib:
Hairer as complete book 

Image_b_direct.tex:
Notation change

Image_b_convex.tex:
Notation change, 
Make it fit into context of direct adaptation

\title{Positivity-Preserving Adaptive Runge-Kutta Methods}
\author{Stephan Nüßlein}

@@

Many physical processes can be described with differential equations. 
The physical property that are involved in these processes often only make sense if they are positive or non-negative. 
Concentrations or number of particles have to be $\geq 0$.
Other values may only make sense if they comply with other restrictions. An Example of these would-be probabilities or fractions which have to be $\geq 0$ and $\leq 1$.
The ordinary differential equations (ODE) or a partial differential equations (PDEs) are often too complex to solve them analytically and therefore require numerical methods to get an approximate solution.
Numerical methods generally do not satisfy these restrictions on the solutions.

The most common methods to solve ODEs are Runge–Kutta methods (RKMs). 
With the method of lines (MOL) it is also possible to solve PDEs with RKMs.
At first the PDE is discretized in space. The resulting semi-discrete problem is an ODE that can be solved with a RKM. 
It is possible to semi-discretize PDEs in a ways that preserve positivity\cite{kopecz_comparison_2019,kopecz_unconditionally_2018}.

Therefore, it is interesting to know how to use RKMs that preserve positivity.
For RKMs some theoretical limitations are known. One is that any unconditionally positive RKM has an order $\leq 1$. \cite{hundsdorfer_numerical_2003,bolley_conservation_1978,horvath_positivity_1998}. The only well know RKM that is unconditionally positive is the backward euler method, which is obliviously of first order. 
This causes restrictions on the step size for any RKM with a higher order.

There are multiple approaches to get positivity preserving numerical schemes.
One method is the Strong Stability Preserving Runge Kutta (SSPRK). 
The idea is to construct a RKM that can be written as convex combination of euler steps. 
By this the stability properties of the forward euler step are carried over to the SSPRK.
The same happens to the positivity properties. By this the positivity of the method is ensured for time steps $\dt$ with $\dt \leq R \dt_{FE}$, where $R$ depends on the used SSPRK and $\dt_{FE}$ is the forward euler time step \cite{gottlieb_strong_2011}. 
The Modified Patanka Runge Kutta (MPRK) are 
based on the Patanka RKM. In these destruction Terms are scaled by the new solution. This preserves positivity but violates the mass conservation. The MPRK overcome this disadvantage by also scaling the production terms. This is done by solving a linear equation system \cite{kopecz_comparison_2019}.
An approach to ensure positivity in a practical application is shown in \cite{shampine_non-negative_2005}. Ensuring the positivity is done by redefining the ODE outside the domain. Another technique used are event finding methods. This approach is implemented in the Matlab ODE Suite.
Unconditionally positive method of order higher than 1 are the diagonally split Runge-Kutta (DSRK) methods \cite{horvath_positivity_1998}. These are no longer part of the general linear methods. This allows them to be positive and higher order.

In the following paper an approach is taken that does not overcome the order imitation but is based on the idea that in practical problems for many time steps positivity can be achieved with a higher order methods.
The main idea is to adapt the RKM after computation of the stage values in a way that ensures positivity.  
This means that the computed stage values can be used and don't have to be recomputed. With increasing complexity of the right-hand side, a step reject has a high cost. This makes the additional cost of adapting the RKM economically.
The resulting b should comply with the order conditions as far as possible.

A common way to calculate two different solutions with one set of stage values is the use of embedded methods for error approximation. 
The idea of adapting the weights $b$ after calculating the stage values is used in \cite{ketcheson_spatially_2013}
In this case it is used to adapt the properties of the time integrator for a method of lines simulation on a PDE. With this approach, it is possible to have different properties of the RKM at different parts of the domain. 
Another method that adapts the $b$ are the relaxation Runge–Kutta (RRK) methods. 
In these the weights are scaled by a scalar relaxation parameter to conserve convex quantities. This can be used to conserve the energy or entropy \cite{ranocha_relaxation_2019,ketcheson_relaxation_2019}

In this paper, an adaption of the $b$ is done by solving a Linear programming (LP)-Problem.
These are convex optimization problems where all constraints and the objective function are all affine. These can incorporate both equality and inequality constraints. 
LP-Problems can be quite reliably solved with the Simplex method or with interior point methods. This can be done in reasonable time even for large problems. 
\cite{boyd_convex_2004}

The paper unfolds as follows. In section\,\ref{sec:main_idea} the main idea is explained. How the method can be written as LP-Problem is explained in section\,\ref{sec:LP-Problem}.
In section\,\ref{sec:integration} it is described how the new approach can be used with different RKM, how it can be integrated in a step size control and how the Region of absolute stability can be approximated.  
In section\,\ref{sec:imple} further details that were used when implementing the algorithm are described.
In section\,\ref{sec:Numeric_Results} Numerical results are given for multiple test problems.
A conclusion is given in section\,\ref{sec:conclusion}

@@

When computing the solution of an ODE $u ' = f(t,u) $ using a RKM with $s$ Stages and the Butcher tableau
\begin{align}
\begin{array}{c|c}
c &  A \\
\hline
 & b\\
\end{array}
\end{align}
the stage values $y^{(1)},\cdots,y^{(s)}$ are computed according to
\begin{equation}
y^{(j)} =  u^n + \dt \sum_{k = 1}^{s} a_{jk} f(t^n + \dt c_k,y^{(k)})  \quad j = 1,\cdots,s
\end{equation}
Based on these values the next solution $u^{n+1}$ is computed by
\begin{equation}
u^{n+1} = u^n + \dt \sum_{j  = 1}^s f(t^n + \dt c_j,y^{(j)}) b_j .
\end{equation}
The stage derivatives $f^{(1)},\cdots,f^{(s)}$ can be calculated with
\begin{equation}
f^{(j)} = f(t^n + \dt c_j,y^{(j)})
\end{equation}
The new solution can now be calculated with
\begin{equation}\label{eq:Combination}
u^{n+1} = u^n + \dt F b
\end{equation}
where $F$ is a matrix containing the stage derivatives $f^{(1)},\cdots,f^{(s)}$.

The new solution $u^{n+1}$ does not comply with positivity for some cases. 
So it may be advantageous to adapt \eqref{eq:Combination} in such a way that $u^{n+1}$ is positive, instead of rejecting the Solution and computing a new set of stage values. This is done by adapting the weights $b$.

For the computation of the stage derivatives only the $A$ has to be known. The $b$ is not needed for this computation. 
Based on this Observation we can define $RKM_{A}$ as the set of RKMs that share the same $A$ and only differ in $b$.
Calculating the stage derivatives can be considered as solving the ODE for all RKM in $RKM_A$ at the same time.
The idea now is to choose one $RKM_{(A,b)} \in RKM_{A}$ by fixing the $b$. 
Because the stage derivatives are already known we can choose a $RKM_{(A,b)}$ that leads to a positive $u^{n+1}$, if $RKM_{(A,b)} \in RKM_{A} | u^{n+1} \geq 0$ exists.
When altering the $b$ the properties of an RKM are preserved. 
Especially this approach preserves linear invariants like mass conservation, because it is a regular RKM.

As it can be see in \eqref{eq:Combination} $u^{n+1}$ is a linear combination of the stage values and the previous solution $u^n$.
Because the coefficients $a_{jl}$ are already fixed the Order Conditions reduce to linear equations.
This means that finding the $b$ can be done by solving a linear Programming problem.

If there are no negative values in $u^{n+1}$ with the original $b$, then the method does not cause additional computational effort, because it only has to solve the LP-Problem if the regular RKM does not lead to a positive $u^{n+1}$. %(except for the test)

@@

To illustrate the use of the method we consider the following reaction equation from\,\cite{kopecz_comparison_2019}.

\begin{subequations}
\label{eq:Reaction}
\begin{align}
u_1' &= 0.01u_2 + 0.01 u_3 +0.003u_4 - \frac{u_1 u_2}{0.01+u_1} \\ 
u_2' &= \frac{u_1u_2}{0.01+u_1}-0.01 u_2-0.5(1-\exp(-1.21 u_2^2)) u_3 -0.05 u_2 \\ 
u_3' &= 0.5(1-\exp(-1.21u_2^2)) u_3 - 0.01 u_3 -0.02 u_3 \\ 
u_4' &=0.05 u_2 + 0.02 u_3 0.003u_4 
\end{align}
\end{subequations}
and the initial conditions
\begin{equation}
u(0) = (8,2,1,4)^T
\end{equation}
When solved with the Cash-Karp (CK5) method and $\dt = 0.005$ the approximated solution contains negative values. This causes qualitatively wrong solutions to the problem. 
In Figure\,\ref{fig:exampleI} the obtained results are plotted with a dashed line. 
At $t=1.905$ $u_1$ gets negative. This leads to a diverging solution.

Now the $b$ are adapted. The results are also plotted in Figure\,\ref{fig:exampleI}. 
All $u\geq0$ is now fulfilled. A qualitatively correct solution is obtained.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.75\textwidth]{plots/exampleI.pdf}
    \caption{Numerical approxiamtion of Reaction Problem from \cite{kopecz_comparison_2019}. Computed with ck5 and $\dt = 0.005$. The dashed line are the reference solution without the adaption of the $b$. The lower plot are the weights $b$ }
    \label{fig:exampleI}
\end{figure}

The used weights $b_1,\cdots,b_5$ are also plotted in Figure\,\ref{fig:exampleI}. 
The used weights are equal the original weights for $t<1.905$. At $t=1.905$ the weights are first adapted to ensure the positivity of the solution. For $t>2.63$ the original set of weights lead to a positive solution. Because of this, the used weights equal the original weights again.

\section{Details (TODO: better title)}\label{sec:LP-Problem}

In the following section, it is explained how the approach can be written as a LP-Problem. 
For this we have to consider which $b$ to choose. 
Generally, any $b \in \R$ can be used. 
A first requirement would be that the new solution is positive.
We can further reduce this space by enforcing the order conditions. 
By incorporating the order conditions, the space of $b$ is limited to a subspace of $\R$. %(Note: not linear because 0 is not in subspace)
Finally, we have to give a objective function to get the optimal $b$.

@@

The main goal is to choose a $RKM_{(A,b)}$ that for which $u^{n+1}$ resembles the solution of the ODE $u(t_{n+1})$. 
A obvious objective is to use a $RKM_{(A,b)}$ with a high order, but this is not enough.
To get a better understanding for the method we consider the behavior of a simple problem.

@@

\begin{equation}
u'(t) = L u(t) \qquad L = \left[\begin{matrix}- a & 1\\a & -1\end{matrix}\right] \quad a =5
\end{equation}
and the first order RKM with two stages
\begin{align}
\begin{array}{c|cc}
0 &  & \\
\frac{1}{2} & \frac{1}{2} & \\
\hline
 & \frac{1}{2} & \frac{1}{2}\\
\end{array}
\end{align}
If simulated with $\dt = 1$ the stage derivatives are 
\begin{equation}
f^1 = \left(\begin{matrix}-5\\5\end{matrix}\right) \qquad f^2 = \left(\begin{matrix}10\\-10\end{matrix}\right)
\end{equation}
All $b$ that comply with the order Condition for the first order can be expressed as
\begin{equation}
b= \left(\begin{matrix}\frac{1}{2}\\ \frac{1}{2}\end{matrix}\right) + \alpha \left(\begin{matrix}1\\ -1\end{matrix}\right)
\end{equation}
With this we have one degree of freedom for the choice of $b$. This is done by the parameter $\alpha$.
If the general expression for $b$ is inserted in \eqref{eq:Combination} the general solution is 
\begin{equation}
u^{1} = u^0 + \dt  \left[f^1,f^2\right] b = \left(\begin{matrix}\frac{5}{2}\\- \frac{5}{2}\end{matrix}\right) \alpha  \left(\begin{matrix}-15\\15\end{matrix}\right)
\end{equation}
The solution for the original RKM would be $u^{1} = \left(\frac{5}{2} ,- \frac{5}{2}\right)^T$ which contains a negative value. But by changing the parameter $\alpha$ the weights $b$ and the solution is altered.
By a suitable choice of $\alpha \in \left[\frac{1}{6},\frac{7}{30}\right]$ any $u$ that complies with mass conservation and positivity can be reached. 
By adding additional constraints on the $b$ the choice of $\alpha$ can be narrowed down. 
An objective function is also needed to make the choice unique. This should be designed in a way to prefer $b$ that are close to the original weights $b_{orig}$

@@

The order conditions for a RKM are a set of Equations depending on $a_{jl}$, $c_1,\cdots,c_s$ and $b_1,\cdots,b_s$. In these Equations $b_1,\cdots,b_s$ only appear linearly \cite{hairer_runge-kutta_1993}.
If $a_{jl}$ and $c_1,\cdots,c_s$ are already known, the order conditions for the Order $p$ are an linear equations system and can be written as $Q b = r$.

@@

If the $A$ of a preexisting RK-method is used, there is at least one solution for the linear equation system.
In order to get an optimization problem, the equation system has to be underdetermined.
This implies that $\mathrm{rank}(Q) < s$. 
Because the Quadrature conditions are linear independent, there are at least $p$ independent vectors in $Q$. This leads to the condition $p < s$.
This implies that the number of stages has to be higher than the wanted order.

@@

There are two possible ways to write down the adaptation of the $b$. The first way is to directly adapt the original weights $b_{orig}$ by adding a correction $\Delta b$ that ensures positivity.
This is the easiest way to adapt the $b$. An advantageous property is that the step size can be reduced after calculation the stage values by using the dense output order conditions.  
This approach suits best for explicit methods.
The direct adaption approach is explained in more detail in section\,\ref{sec:direct}.

The second approach is to choose $b$ as a convex combination of a set of $b$s. This approach is more useful if some embedded methods with certain properties are known. 
This approach also shows more predictable behavior because the resulting $u^{n+1}_b$ is also a convex combination of the solutions $u^{n+1}_{b_a},u^{n+1}_{b_b},\cdots$ 
This approach works best with implicit methods.
The direct adaption approach is explained in more detail in section\,\ref{sec:convex}

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_direct}
        \caption{Direct Adaptation}
        \label{fig:b_direct}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_convex}
        \caption{Convex Combiantion}
        \label{fig:b_convex}
    \end{subfigure}
    \caption{Graphical representation of the two different approaches to adapt the weights}\label{fig:b_space}
\end{figure}

@@

When directly adapting the $b$ a new $b$ that complies with the order conditions is chosen.   
The weights $b = b_{orig} + \Delta b$ consist of the weights of the original method $b_{orig}$ and an adaption $\Delta b$. The adaption is $\Delta b \in \mathrm{ker}(Q)$. 
In Figure\,\ref{fig:b_direct} this is drawn. The red line represents the subspace of $b$ that satisfy the order conditions. Both $b_{orig}$ and $b$ are in this subspace.

To run it as a LP-Problem a set of constraints is constructed.  
Also, an objective function is needed to determine the optimal $b$.

\subsubsection{Constraints}
The Direct Adaptation approach leads to two set of constraints on the $b$.
The first set of constraints is the positivity and can be written as 
\begin{equation}
u_{n+1}=u_n+\dt K \vec{b} \geq 0
\end{equation}
The second set of constraints are the order conditions
\begin{equation}
Qb=r
\end{equation}

@@

A reasonable property of the objective function would be that for the LP-Problem only constrained by the order condition the optimal solution is $b = b_{orig}$. 
This can be achieved by $argmin(f_{optim}(b)) = b_{orig}$. 
In practice, this also means that there is no need to solve solve the optimization Problem if no constraints are violated because the solution of this problem is already known. This is advantageous for performance reasons.
The different approaches have been tested on a linear test problem from \cite{kopecz_unconditionally_2018}. The optimization problem was used at every step.
The magnitude of $b$ can get large if the order conditions are used as objective function.
Large values of b can lead to numerical problems. Because $\sum_{i  = 1}^s b_i = 1$, an increase of $||b||$ leads to a combination of negative and positive values. Because the computation is done with a finite precision (and usually in floating point) large $||b||$ can lead to absorption effects.    
By changing the $b$ the stability function is altered. This can lead to oscillations if the poles of the RHS are no longer in the stability region of the resulting method.
For these reasons $||b-b_{orig}||$ seems like a good choice. The 1-Norm can be implement the with Linear Programming using slack variables. We choose

@@

If the original step $\dt$ was too large and there is no suitable $b$, we would like to calculate a $u^{n+1}$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the order conditions for dense Output. These can be found in \cite{hairer_runge-kutta_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \dt$ whereas $\dt$ is the time step used when calculating the stage values.
To calculate a reduced step the order conditions have to be adapted and a new $b_{orig}$ for the objective function (and more important the Error detection afterwards) has to be generated. 
(TODO: the $b_{orig}$ is still a open question)

When adapting the Order Condition to a new $\theta$ the Order Condition Matrix $Q$ stays unchanged whereas the right hand side depends on $\theta$.

The dependency of $r(\theta)$ is nonlinear. This means that the maximum $\theta$ with a suitable $b$ cannot be calculated by solving a LP-Problem. This means that different $\theta$ have to be tested. 
(Note: we could include some algorithm to search for it)

\subsection{Convex combination of existing methods}\label{sec:convex}
For some test problems (e.g. the Heat Equation solved with an implicit method of higher order), large distortions occur, even though a norm of the distortion of the b is used as an objective function. 
For implicit methods, it is known that the backward euler steps lead to a positive solution for any $\dt$ \cite{hundsdorfer_numerical_2003}. The backward Euler is still only first Order but shows more exact results than a method of higher Order that is distorted by the adaption of the $b$, at least at some steps. 
A possible method to fix this is to generate the new b based on a convex combination of embedded methods of RKM.

>>>>

\title{Positivity-Preserving Adaptive Runge-Kutta Methods}
\author{Nüßlein, Stephan and Ketcheson, David I.}

@@

Many physical processes can be described with differential equations. 
The physical quantities that are involved in these processes often only make sense if they are positive or non-negative. 
Concentrations or the number of particles have to be $\geq 0$.
Other values may only make sense if they comply with other restrictions. An example of these would be probabilities or fractions which have to be $\geq 0$ and $\leq 1$.
The ordinary differential equations (ODE) or partial differential equations (PDEs) are often too complex to solve them analytically and therefore require numerical methods to get an approximate solution.
Numerical methods generally do not satisfy these restrictions on the solutions.

The most common methods to solve ODEs are Runge–Kutta methods (RKMs). 
With the method of lines (MOL) it is also possible to solve PDEs with RKMs.
At first the PDE is discretized in space. The resulting semi-discrete problem is an ODE that can be solved with a RKM. 
It is possible to semi-discretize PDEs in ways that preserve positivity\cite{kopecz_comparison_2019,kopecz_unconditionally_2018}.

Therefore, it is interesting to know how to use RKMs that preserve positivity.
For RKMs some theoretical limitations are known. One is that any unconditionally positive RKM has an order $\leq 1$. \cite{hundsdorfer_numerical_2003,bolley_conservation_1978,horvath_positivity_1998}. The only well-known RKM that is unconditionally positive is the backward euler method, which is obviously of first order. 
This causes restrictions on the step size for any RKM with a higher order.

There are multiple approaches to get positivity preserving numerical schemes.
One method is the Strong Stability Preserving Runge Kutta (SSPRK). 
The idea is to construct a RKM that can be written as convex combination of euler steps. 
By this the stability properties of the forward euler step are carried over to the SSPRK.
The same happens to the positivity properties. By this the positivity of the method is ensured for time steps $\dt$ with $\dt \leq R \dt_{FE}$, where $R$ depends on the used SSPRK and $\dt_{FE}$ is the forward Euler time step \cite{gottlieb_strong_2011}. 
The Modified Patankar Runge Kutta (MPRK) are 
based on the Patankar RKM. In these destruction Terms are scaled by the new solution. This preserves positivity but violates the mass conservation. The MPRK overcome this disadvantage by also scaling the production terms. This is done by solving a linear equation system \cite{kopecz_comparison_2019}.
An approach to ensure positivity in a practical application is shown in \cite{shampine_non-negative_2005}. Ensuring the positivity is done by redefining the ODE outside the domain. Another technique used is event finding methods. This approach is implemented in the Matlab ODE Suite.
Unconditionally positive method of order higher than 1 are the diagonally split Runge-Kutta (DSRK) methods \cite{horvath_positivity_1998}. These are no longer part of the general linear methods. This allows them to be positive and higher order.

In the following paper an approach is taken that does not overcome the order limitation but is based on the idea that in practical problems for many time steps positivity can be achieved with higher order methods.
The main idea is to adapt the RKM after the computation of the stage values in a way that ensures positivity.  
This means that the computed stage values can be used and don't have to be recomputed. With increasing complexity of the right-hand side, a step reject has a high cost. This makes the additional cost of adapting the RKM economical.
The resulting weights should comply with the order conditions as far as possible.

A common way to calculate two different solutions with one set of stage values is the usage of embedded methods for error approximation. 
The idea of adapting the weights after calculating the stage values is used in \cite{ketcheson_spatially_2013}.
In this case it is used to adapt the properties of the time integrator for a method of lines solution of a PDE. With this approach, it is possible to have different properties of the RKM at different parts of the domain. 
Other methods that adapts the weights are the relaxation Runge–Kutta (RRK) methods. 
In these, the weights are scaled by a scalar relaxation parameter to conserve convex quantities. This can be used to conserve the energy or entropy \cite{ranocha_relaxation_2019,ketcheson_relaxation_2019}

In this paper, an adaptation of the weights is done by solving a Linear program (LP).
These are convex optimization problems where all constraints and the objective function are affine. These can incorporate both equality and inequality constraints. 
LPs can be quite reliably solved with the Simplex method or with interior point methods. This can be done in reasonable time even for large problems. 
\cite{boyd_convex_2004}

The paper unfolds as follows. In section\,\ref{sec:main_idea} the main idea is explained. How the method can be written as LP is explained in section\,\ref{sec:LP}.
Section\,\ref{sec:integration} describes how the new approach can be used with different RKM, how it can be integrated in a step size control and how the Region of absolute stability can be approximated.  
In section\,\ref{sec:imple} further details that were used when implementing the algorithm are described.
In section\,\ref{sec:Numeric_Results} Numerical results are given for multiple test problems.
A conclusion is given in section\,\ref{sec:conclusion}.

@@

When computing the solution of an ODE $u ' = f(t,u) $ using a RKM with $s$ Stages and the Butcher tableau
\begin{align}
\begin{array}{c|c}
c &  A \\
\hline
 & b\\
\end{array}
\end{align}
the stage values $y_1,\cdots,y_s$ are computed according to
\begin{equation}
y_j =  u^n + \dt \sum_{k = 1}^{s} a_{jk} f(t^n + \dt c_k,y_k)  \quad j = 1,\cdots,s
\end{equation}
Based on these values the next solution $u^{n+1}$ is computed by
\begin{equation}
u^{n+1} = u^n + \dt \sum_{j  = 1}^s f(t^n + \dt c_j,y_j) b_j .
\end{equation}
The stage derivatives $f_1,\cdots,f_s$ can be calculated with
\begin{equation}
f_j = f(t^n + \dt c_j,y_j)
\end{equation}
The new solution can now be calculated with
\begin{equation}\label{eq:Combination}
u^{n+1} = u^n + \dt F b
\end{equation}
where $F$ is a matrix containing the stage derivatives $F = \big[f(y_1) , \cdots f(y_s)\big]$.

The new solution $u^{n+1}$ does not comply with positivity for some cases. 
So it might be advantageous to adapt \eqref{eq:Combination} in such a way that $u^{n+1}$ is positive, instead of rejecting the solution and computing a new set of stage values. This is done by adapting the weights.

For the computation of the stage derivatives only the $A$ has to be known. The weights is not needed for this computation. 
Based on this observation we can define $(A,\bullet)$ as the set of RKMs that share the same $A$ and only differ in the weights.
Calculating the stage derivatives can be considered as solving the ODE for all RKM in $(A,\bullet)$ at the same time.
The idea is to choose one $(A,b) \in (A,\bullet)$ by fixing the $b$. 
Because the stage derivatives are already known we can choose a $(A,b)$ that leads to a positive $u^{n+1}$, if $(A,b) \in (A,\bullet) | u^{n+1} \geq 0$ exists.
When altering the weights the properties of a RKM are preserved. 
Especially this approach preserves linear invariants like mass conservation, because it is a regular RKM.

As it can be seen in \eqref{eq:Combination} $u^{n+1}$ is a linear combination of the stage values and the previous solution $u^n$.
Because the coefficients $A$ are already fixed the Order Conditions reduce to linear equations.
This means that finding the weights can be done by solving a linear Programming problem.

If there are no negative values in $u^{n+1}$ with the original weights, the method does not cause additional computational effort, because it only has to solve the LP if the regular RKM does not lead to a positive $u^{n+1}$. %(except for the test)

@@

To illustrate the usage of the method we consider the following reaction equation from\,\cite{kopecz_comparison_2019}.

\begin{subequations}
\label{eq:Reaction}
\begin{align}
u_1' &= 0.01u_2 + 0.01 u_3 +0.003u_4 - \frac{u_1 u_2}{0.01+u_1} \\ 
u_2' &= \frac{u_1u_2}{0.01+u_1}-0.01 u_2-0.5(1-\exp(-1.21 u_2^2)) u_3 -0.05 u_2 \\ 
u_3' &= 0.5(1-\exp(-1.21u_2^2)) u_3 - 0.01 u_3 -0.02 u_3 \\ 
u_4' &=0.05 u_2 + 0.02 u_3 + 0.003u_4 
\end{align}
\end{subequations}
and the initial conditions
\begin{equation}
u(0) = (8,2,1,4)^T
\end{equation}
When solved with the Cash-Karp (CK5) method and $\dt = 0.005$ the approximated solution contains negative values. This causes qualitatively wrong solutions to the problem. 
In Figure\,\ref{fig:exampleI} the obtained results are plotted with a dashed line. 
At $t=1.905$ $u_1$ gets negative. This leads to a diverging solution.

Now the weights are adapted. The results are also plotted in Figure\,\ref{fig:exampleI}. 
All $u\geq0$ is now fulfilled. A qualitatively correct solution is obtained.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.75\textwidth]{plots/exampleI.pdf}
    \caption{Numerical approximation of the Reaction Problem from \cite{kopecz_comparison_2019}. Computed with ck5 and $\dt = 0.005$. The dashed lines are the reference (Note: Reference might be misleading here because real solution is different) solution without the adaption of the weights. The lower plot are the weights }
    \label{fig:exampleI}
\end{figure}

The used weights $b_1,\cdots,b_5$ are also plotted in Figure\,\ref{fig:exampleI}. 
The used weights are equal to the original weights for $t<1.905$. At $t=1.905$ the weights are first adapted to ensure the positivity of the solution. For $t>2.63$ the original set of weights lead to a positive solution. Because of this, the used weights equal the original weights again.

\section{Details (TODO: better title)}\label{sec:LP}

In the following section, it is explained how the approach can be written as a LP. 
For this we have to consider which weights to choose. 
Generally, any $b \in \R$ can be used. 
A first requirement would be that the new solution is positive.
We can further reduce this space by enforcing the order conditions. 
By incorporating the order conditions, the space of possible weights is limited to a subspace of $\R$. %(Note: not linear because 0 is not in subspace)
Finally, we have to give an objective function to get the optimal weights.

@@

The main goal is to choose a $(A,b)$ in such a way that $u^{n+1}$ resembles the solution of the ODE $u(t_{n+1})$. 
A obvious objective is to use a $(A,b)$ with a high order, but this is not enough.
To get a better understanding for the method we consider the behavior of a simple problem.

@@

\begin{equation}
u'(t) = L u(t) \qquad L = \left[\begin{matrix}- 5 & 1\\5 & -1\end{matrix}\right] 
\end{equation}
and and use the three Stage and third order SSP method (SSP33)
\begin{align}
\begin{array}{c|ccc}
0 &  &  & \\
1 & 1 &  & \\
\frac{1}{2} & \frac{1}{4} & \frac{1}{4} & \\
\hline
 & \frac{1}{6} & \frac{1}{6} & \frac{2}{3}\\
\end{array}
\end{align}
The Matrix $L$ has the eigenvalues $\lambda_1 = -6,\lambda_2=0$. The real stability interval of the SSP33 method includes $[-2.5,0]$. The method is absolut stable for this problem with $\dt = \frac{1}{3}$
The stage derivatives are
\begin{equation}
f(y_1)=\left[\begin{matrix}-5\\5\end{matrix}\right]\qquad f(y_2)=\left[\begin{matrix}5\\-5\end{matrix}\right]\qquad f(y_3)=\left[\begin{matrix}-5\\5\end{matrix}\right]\end{equation}
The values of the next step using the standard weights is
\begin{equation}
u_n=\left[\begin{matrix}- \frac{1}{9}\\\frac{10}{9}\end{matrix}\right]
\end{equation}
The new solution contains negative values. 
Therefor we want to adapt the weights to ensure positifity. 
All weights that comply with the order Condition for the first and second order can be expressed as
\begin{equation}
\tilde b=\left[\begin{matrix}\frac{1}{6}\\\frac{1}{6}\\\frac{2}{3}\end{matrix}\right]+\alpha \left[\begin{matrix}\frac{1}{2}\\\frac{1}{2}\\-1\end{matrix}\right]
\end{equation}
With this we have one degree of freedom for the choice of the weights. This is done by the parameter $\alpha$.
If the general expression for the weights is inserted in \eqref{eq:Combination} the general solution is 
\begin{equation}
u^{1} = u^0 + \dt  \left[f(y_1),f(y_2),f(y_3)\right] b =
\left[\begin{matrix}- \frac{1}{9}\\\frac{10}{9}\end{matrix}\right] +\alpha \left[\begin{matrix}\frac{5}{3}\\- \frac{5}{3}\end{matrix}\right]
\end{equation}
By changing the parameter $\alpha$ the weights $b$ and the solution is altered.
With a suitable choice of $\alpha \in \left[\frac{1}{15},\frac{2}{3}\right]$ any $u$ that complies with mass conservation and positivity can be reached. 
By adding additional constraints on the weights the choice of $\alpha$ can be narrowed down. 
An objective function is also needed to make the choice unique. This should be designed in a way to prefer weights that are close to the original weights.

@@

The order conditions for a RKM are a set of Equations depending on $A$, $c$ and $b$. In these Equations $b$ only appears linearly \cite{hairer_solving_1993}.
If $A$ and $c$ are already known, the order conditions for the Order $p$ are a linear equations system and can be written as $Q b = r$.

@@

If the $A$ of a preexisting RK-method is used, there is at least one solution for the linear equation system.
In order to get an optimization problem, the equation system has to be underdetermined.
This implies that $\mathrm{rank}(Q) < s$. 
Because the quadrature conditions are linearly independent, there are at least $p$ independent vectors in $Q$. This leads to the condition $p < s$.
This implies that the number of stages has to be higher than the wanted order.

@@

There are two possible ways to write down the adaptation of the weights. The first way is to directly adapt the original weights $b$ by adding a correction $\Delta b$ that ensures positivity.
This is the easiest way to adapt the weights. An advantageous property is that the step size can be reduced after calculating the stage values by using the dense output order conditions.  
The direct adaption is explained in more detail in section\,\ref{sec:direct}.

The second way to formulate a LP is to choose weights as a convex combination of a set of weights. This is more useful if some embedded methods with certain properties are known. 
The convex combination also shows more predictable behavior because the resulting $u^{n+1}_b$ is also a convex combination of the solutions $u^{n+1}_{b_a},u^{n+1}_{b_b},\cdots$ 
This is explained in more detail in section\,\ref{sec:convex}

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_direct}
        \caption{Direct Adaptation}
        \label{fig:b_direct}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_convex}
        \caption{Convex Combination}
        \label{fig:b_convex}
    \end{subfigure}
    \caption{Graphical representation of the two different approaches to adapt the weights}\label{fig:b_space}
\end{figure}

@@

When directly adapting the weights new weights that complies with the order conditions is chosen.   
The weights $\tilde{b} = b + \Delta b$ consist of the weights of the original method $b$ and an adaption $\Delta b$. The adaption is $\Delta b \in \mathrm{ker}(Q)$. 
In Figure\,\ref{fig:b_direct} this is drawn. The red line represents the subspace of weights that satisfies the order conditions. Both $b$ and $\tilde{b}$ are in this subspace.

To run it as a LP a set of constraints is constructed.  
Also, an objective function is needed to determine the optimal weights.

\subsubsection{Constraints}
The Direct Adaptation approach leads to two sets of constraints on $b$.
The first set of constraints is the positivity and can be written as 
\begin{equation}
u_{n+1}=u_n+\dt F \tilde{b} \geq 0
\end{equation}
The second set of constraints are the order conditions
\begin{equation}
Qb=r
\end{equation}

@@

A reasonable property of the objective function would be that for the LP only constrained by the order condition the optimal solution is $b = b_{orig}$. 
This can be achieved by $argmin(f_{optim}(b)) = b_{orig}$. 
In practice, this also means that there is no need to solve the optimization problem if no constraints are violated because the solution of this problem is already known. This is advantageous for performance reasons.
The different approaches have been tested on a linear test problem from \cite{kopecz_unconditionally_2018}. The optimization problem was used at every step.
The magnitude of the weights can get large if the order conditions are used as objective function.
Large values of b can lead to numerical problems. Because $\sum_{i  = 1}^s b_i = 1$, an increase of $||b||$ leads to a combination of negative and positive values. Because the computation is done with a finite precision (and usually in floating point) large $||b||$ can lead to absorption effects.    
By changing the weights the stability function is altered. This can lead to oscillations if the poles of the RHS are no longer in the stability region of the resulting method.
For these reasons $||b-b_{orig}||$ seems like a good choice. The 1-Norm can be implemented the with Linear Programming using slack variables. We choose

@@

If the original step $\dt$ was too large and there is no suitable set of weights, the LP is infeasible. In this case we would like to calculate $u^{n+\theta}$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the order conditions for dense output. These can be found in \cite{hairer_solving_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \dt$ whereas $\dt$ is the time step used when calculating the stage values.
To calculate a reduced step the order conditions have to be adapted and a new $b_{orig}$ for the objective function (and more important the error detection afterwards) has to be generated. 
(TODO: the $b_{orig}$ is still a open question)

When adapting the Order Condition to a new $\theta$ the Order Condition Matrix $Q$ stays unchanged whereas the right hand side $r(\theta)$ depends on $\theta$.

The dependency of $r(\theta)$ is nonlinear. This means that the maximum $\theta$ with suitable weights cannot be calculated by solving a LP. This means that different $\theta$ have to be tested. 
(Note: we could include some algorithm to search for it)

\subsection{Convex combination of existing methods}\label{sec:convex}
For some test problems (e.g. the Heat Equation solved with an implicit method of higher order), large changes occur, even though a norm of $\Delta b$ is used as an objective function. A numerical example is given in Section\,\ref{sec:Numeric_Results}
It is known that the backward Euler method lead to a positive solution for any $\dt$ \cite{hundsdorfer_numerical_2003}. The backward Euler is still only first Order but shows more exact results than a method of higher Order that is distorted by the adaption of the weights, at least at some steps. 
A possible method to fix this is to generate the new b based on a convex combination of embedded methods.
